The imprint of global magma oceans on exoplanet demographics

### 2023-02-27 src/tex/ms.tex
draft discussion of missions diagnostic power

\todo[inline]{ARNAUD: revise definition of $S$ following our discussion on 2022-12-22. Introduce net instellation and clarify what is meant by it.}
\begin{note}
    The power from the host star per unit area at the position of a planet or net instellation $S$ in units of Earth's insolation is given by
    \begin{equation}
        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
    \end{equation}
    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's [CITE].
    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
%    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
   We do not take into account additional heating sources such as tidal effects~\citep[e.g.][]{Barnes2013}.
\end{note}

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical detection of the runaway greenhouse threshold.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

%We consider a number of key parameters of transit surveys: The telescope's mirror diameter (or effective diameter for a telescope array) $D$, the total time budget of the survey for observations $t_\mathrm{total}$, the maximum number of observed transits for each target $N_\mathrm{obs, max}$, the slew time between observations $t_\mathrm{slew}$, ... \todo[]{double-check which survey params are needed}
We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
\begin{itemize}
    \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas. This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
    \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

@@

\subsection{Diagnostic power of upcoming exoplanet missions}\todo[inline]{\plato: needed sample size, precision}
The fidelity of a future detection or falsification of a magma ocean signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection given the existence of the effect.
...

The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
...

@@

%\begin{acknowledgments}
%The authors thank Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
%\end{acknowledgments}

>>>>

\todo[inline]{ARNAUD: revise definition of $S$ following our discussion on 2022-12-22. Introduce net instellation and clarify what is meant by it.}
\begin{note}
    The power from the host star per unit area at the position of a planet or net instellation $S$ in units of Earth's insolation is given by
    \begin{equation}
        \frac{S}{S_\oplus} = \left(\frac{L_\star}{L_\odot}\right) \left(\frac{au}{a}\right)^2 .
    \end{equation}
    Here, we assume global redistribution of incoming flux and a fixed albedo of $0.3$, comparable to Earth's~\citep{Haar1971}.
    The absolute value of $S_\oplus$, by which all synthetic planets are scaled, is then \SI{238}{\watt\per\square\meter}.
%    We further define the solar-equivalent semi-major axis $a_{eff} = a (L_\star/L_\odot)^{-1/2}$, at which a planet experiences the same instellation as a Solar System planet at an orbital distance~$a$.
   We do not take into account additional heating sources such as tidal effects~\citep[e.g.][]{Barnes2013}.
\end{note}

@@

\subsection{Testability of the runaway greenhouse hypothesis}\label{sec:res_testability}
\begin{figure}[ht!]
%    \script{figurescript.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/optimistic_R-S.pdf}
        \caption{
        Detection of the runaway greenhouse threshold in the optimistic case.
        From simulated radius and instellation measurements of a large ($N = \var{N_optimistic}$) survey, we compute the moving average (blue confidence intervals) and fit the runaway greenhouse hypothesis to it (Eqn.~\ref{eq:rgh_hypo}, random draws from the posterior in red).
            The pattern is detected with high significance.
        }
        \label{fig:optimistic_R-S}
    \end{centering}
\end{figure}
Figure~\ref{fig:optimistic_R-S} shows a prototypical statistical detection of the runaway greenhouse transition.
For the sake of clarity, we chose an optimistic scenario where the sample is large ($N = \var{N_optimistic}$) and the measurement uncertainties are small ($\sigma_{R} = \sigmaR, \sigma_{S} = \sigmaS$).
We assumed that the fraction of those planets irradiated stronger than $S_\mathrm{thresh}$ that have runaway greenhouse climates is \frgh, and we chose a water mass fraction of \wrr\ for each planet.
In this case, the pattern was detected with high significance ($\Delta \ln \mathcal{Z} \approx 100$).

@@

\subsubsection{Key factors influencing tests of the runaway greenhouse hypothesis}
We identified the following drivers of the diagnostic power for detecting the runaway greenhouse transition with a transit survey:
\begin{itemize}
    \item Occurrence rate of planets forming steam atmospheres: The runaway greenhouse climate relies on the production of large amounts of water vapor that then acts as a greenhouse gas. This process works only on planets with an atmosphere and of sufficiently volatile-rich composition\todo[inline]{TIM: how much?}. The fraction of planets fulfilling these requirements has an impact on the amplitude of the demographic imprint of the runaway greenhouse transition.
    \item Planetary evolution and duration of the steam atmosphere phase: An inflated steam atmosphere can only be sustained until the planet has lost its water. Depending on host star spectral type, planetary mass, and composition, planets can spend from a few Myr to several Gyr in runaway greenhouse climates~\citep[][]{Hamano2015,Luger2015}, and only planets observed during this phase will contribute to the signal.
    \item Prevalent water inventory: The magnitude of radius change at the runaway greenhouse threshold is highly sensitive on water mass fraction. As a result, the statistical abundance of water in terrestrial planets impacts the strength of the demographic pattern.
    \item Sample size: Since we are looking for a population-level trend, the statistical significance increases with a larger sample size.
    \item Radius measurement precision: The more precise individual planet radii can be determined, the less smeared out the pattern will be. Good \textit{accuracy} is less important, as long as it does not have a systematic error scaling with stellar irradiance.
    \item Availability and precision of mass measurements: For simple geometric reasons ($\rho \propto R^{-3}$), the expected trend is stronger when measured in bulk density than it is in planet radius space. If transiting planets can be followed up to obtain mass measurements, the statistical significance increases.
\end{itemize}
Besides these main factors, uncertainties in the measured instellations can influence the result, although they are typically small due to the very precise orbital period measurements available for transiting planets.
This can be different for young host stars when their ages can not be well constrained; in particular the long pre-main sequence phase of M~dwarfs shows a large variation in bolometric luminosity~(see Fig.~\ref{fig:luminosity_tracks}).

@@

\subsection{Diagnostic power of near-future exoplanet missions}\label{sec:dis_samplesize}
The fidelity of a future detection or falsification of a runaway greenhouse signature will depend on the significance with which the null hypothesis can be excluded.
As a consequence, instrumentation and survey strategy aiming at testing the hypothesis should aim at maximizing the probability of a true positive detection.
\begin{note}
In Sect.~\ref{sec:res_samplesize} we explore the diagnostic power of space missions that are currently flying or in the planning stage.
We focus here on transit survey missions, as they provide the most direct radius measurements of a practical planet sample.

As discussed above, the key drivers from a mission design perspective are sample size and photometric precision.
We found that along these axes, \plato\ will be the most favorable among the upcoming transit missions.
%The required sample size and precision are well in reach of near-future exoplanet missions such as PLATO.
With an expected transit radius precision of \SI{3}{\percent}~\citep{plato2017} and a yield of $> 100$ planets characterized with such high precision~\citep{Rauer2021}, \plato\ is comparable to the optimistic survey we introduced in Sect.~\ref{sec:res_testability}.
If successful, it should readily detect the predicted statistical imprint or, in case of a non-detection, provide strong upper limits on the occurrence rate of runaway greenhouse planets.

The latter depends on the lifetimes of runaway greenhouse phases, which are a function of the initial water inventory of the planets~\citep{Hamano2015}.
It seems thus feasible to derive the typical water content of rocky planets from these occurrence estimates.
\end{note}

@@

%\begin{acknowledgments}
%The authors thank Gabriele Cugno, Yann Fabel, Paul Molliere, Gijs Mulders, Matthew Murphy, and Terry-Ann Suer for insightful discussions.
%We are grateful to Martin Turbet for providing the mass-radius relationship data for planets harboring steam atmospheres.
%\end{acknowledgments}
